A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database

### 2022-06-08 paper/paper.Rmd
various minor changes for final revision

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques. An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].

@@

For statistics, and data science by association, it is increasingly important to reach the individual. One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual. For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous. Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics. Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak. The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable. However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different. It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility. It would be more useful for an individual to know what percentage of the population have profiles like theirs. We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation. <!--this information is the appropriate message for an individual is that you as an individual are far more than your demographic. The majority of people in the study do not have a pattern that is similar to the overall trend. If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage]. Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.

@@

This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by @HuebnerMariannePhD2016Asat. They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Clean data often then goes through an "initial data analysis" (IDA) [@Chatfield1985TIEo], where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling. This stage is related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data. EDA can be considered to encompass IDA. In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing. @DasuTamraparni2003Edma say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.

Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr]. Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

1. ID: the respondents' ID. 
2. EXPER: stands for experience, the temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
3. LNW: natural logarithm of wages, adjusted with 1990's inflation rate. 
4. BLACK: binary variable, 1 indicates Black and 0 otherwise. 
5. HISPANIC: binary variable, 1 indicates Hispanic and 0 otherwise. 
6. HGC: the highest grade completed.
7. UERATE: the unemployment rate of the year of the survey. When missing, the variable is set to be 7.875 (the average rate).

@@

In the context of official statistics, @LooMarkvander2018Sdcw describe the "statistical value chain," which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the `tidyverse` suite of packages [@tidyverse]. We use `pivot_longer` in `tidyr` [@tidyr] to pivot the data into long-form. For example, we use it to pull the hourly wage (`HRPjob_year`), which is stored in a column corresponding to each job and each year of survey, into columns: job and rate per hour. After that, `separate` from `tidyr` is also used to get the job and survey year in two separate columns. Further `dplyr` [@dplyr] is used and `stringr` [@stringr] are used to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years (as an integers), e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).

For `sex` and `race`, we have simplified the original names provided in the raw data, `SAMPLE_SEX_1979` and `SAMPLE_RACE_78SCRN`, respectively. However, it is important to note that the use of "sex" is probably not correct. This information is provided by the individual, and although it only has two categories, it is more consistent with "gender" as defined in @SAGER. When using the dataset in the classroom, the educators might include discourse on the use of terms "sex" and "gender". Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification. From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.  @kennedy2020using provides a helpful discussion on these topics. <!--Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender. -->

@@

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, encoded as a factor and integer variable, respectively. For each individual, there is only one value of `hgc` and `hgc_i`. This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on). We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types say. Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.

The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as `grade`). This value can change over time, but it is non-decreasing. This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals. Hence, different from `hgc`, this variable could be used to demonstrate longitudinal and time series analysis in the classroom. This is recorded in `new_data_qnames` as columns beginning with `Q3-4` and `HGC` with year as a suffix. In addition, it is also in the columns beginning with `HGCREV` reflecting revised data. We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked. In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly. <!--We also check the consistency between `hgc_i` and `grade`. Some problematic observations were found, for example, `grade` value is greater than `hgc_i`. We then fix this problem by replacing those `grade` that is greater than its `hgc_i`. -->

@@

Our target for the employment-related variables is to obtain the respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (in years) in the workforce (`yr_wforce`), and work experience measured as the number of years worked (`exp`). As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.

@@

For the year of individual started working (`stwork`),  we only rename the column names, which is `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`, from `new_data_qnames`. This variable is then used to calculate the next variable, the length of time in the workforce (`yr_wforce`) for each survey round, which is the year of survey (`year`) minus `stwork`. Finally, experience (`exp`) variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`. To obtain the work experience since 1979, we calculate the cumulative value. As the measurement unit is weeks, we convert this to fractional year.

The employment and demographic variables are then merged into one dataset as in the original data. These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in @SingerJudithD2003Alda. We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data. However, it is worth noting that the original data does not restrict the number of observations for each individual, i.e., there are individuals with only one and two observations.

Finally, we save the resultant wage data on this cohort as `wages`. Note that we save `grade` variable in this dataset instead of `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.

@@

Experience is defined as years after entering the labor force. It represents the difference between the day an individual enters the labor force (`stwork`) relative to the date of the survey, which we call `yr_wforce`. However, using this calculation produced numbers that do not quite match the original data.

@@

In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides. The first step is validating numerical summaries of the raw data are the same as reported by NLSY79. This is followed by graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].

@@

Figure \ref{fig:feature-plot} shows an alternative way to check the data quality. Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals, mostly around 1990. Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals. This tells us there are a number of individuals with unbelievably high maximum wages. Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour. This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.

@@

The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers. It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time. To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed. Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined. Overly smooth profiles would indicate the replacement of values was too severe, resulting in removing the very interesting volatility of wages seen in many individuals. This was achieved using a purpose written `shiny`  [@shiny] app (provided with the code of this paper). <!--A threshold of 0.12 was chosen. That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.--> Observations where a change was made are tracked with a new variable called `is_pred`, so that this effect in the downstream analyses can be monitored.

@@

Figure \ref{fig:fixed-feature-plot} shows the summary statistics after removing extremes. The highest wage overall is now around $1000. Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under $300, and there has been a steady increase in wages over the years. Plot (B) shows there are still a small number of individuals with high maximum wages. Plot (C) shows the profile for ID=39 after imputing the extreme value. The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.

@@

  1. `id`: A unique individual's ID number. This is the `key` of the data as we saved the data as a `tsibble` object. 
  2. `year`: The year the observation was taken. This is the `index` of the data.
  3. `wage`: The mean of the hourly wages the individual gets at each of their different jobs. The value could be a weighted or an arithmetic mean. The weighted mean is used when the information of hours of work as the weight is available. The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
  4. `age_1979`: The age of the individual in 1979.
  5. `sex`: Sex of the individual, f = Female and m = Male.
  6. `race`: Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
  7. `grade`: Integer value of the highest grade completed corresponding to the survey year. 
  8. `hgc`: The highest grade completed ever.
  9. `hgc_i`: Integer value of the highest grade completed ever.
  10. `hgc_1979`: The highest grade completed in 1979 (integer value).

@@

1. `r nrow(join_more_17)` individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset. Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later. 
2. `r nrow(join_less_17)` individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
    i. `r nrow(join_not_elig)` of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
    ii. The information about `ged` is missing in `r nrow(join_missing_ged)` individuals. We decided to include them in the refreshed data. 
    iii. `r nrow(join_ged_both)` individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
    iv. `r nrow(join_not_exist)` individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.

@@

```{r compare-subsets, fig.cap = "Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages. Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 4, out.width = "100%", cache = FALSE}
```

@@

The reporting of highest grade completed is confusing. There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year. To match the original data, it is appropriate to use the `hgc` while in high school. The documentation suggests this variable is available, but it is actually not. Hence, to match the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value. The result does not exactly match the original for a few individuals.

In the original dataset, wages were inflation-adjusted to 1990 prices. This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey. Instead, we have provided a function called `adj_inflation` in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyze the data.

@@

Ultimately, the refreshed data is reasonably similar to the original but disappointingly far from it. The last step required would be to inflation-adjust wages. This is better to do with each wave of new data added so that it is relative to the last date in the data. Our decision was to provide the raw wages and include code to make the adjustment as part of the package.

@@

- It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges. Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
- Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current for today's students.
- It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience. The vignettes of the `brolgar` [@brolgar] package provide useful guides for the case study content.

>>>>

Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples on how to apply different techniques. An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models. The data records hourly wages of a sample of high school dropouts from 1979-1994, along with demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].

@@

For statistics, and data science by association, it is increasingly important to reach the individual. One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public. The two are not the same. As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual. For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous. Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics. Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak. The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable. However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different. It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility. It would be more useful for an individual to know what percentage of the population have profiles like theirs. We argue that different summaries are needed to communicate with the public, so they might be able to individually act on to change their own situation. <!--this information is the appropriate message for an individual is that you as an individual are far more than your demographic. The majority of people in the study do not have a pattern that is similar to the overall trend. If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage]. Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.

@@

This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by @HuebnerMariannePhD2016Asat. They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building. Clean data often then goes through an "initial data analysis" (IDA) [@Chatfield1985TIEo], where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling. This stage is related to exploratory data analysis (EDA), coined by @tukey, with a focus on learning from data. EDA can be considered to encompass IDA. In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing. @DasuTamraparni2003Edma say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.

Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organized into "tidy data" [@WickhamHadley2014TD], and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr]. Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics. The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.

@@

1. ID: the respondents' ID. 
2. EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
3. LNW: natural logarithm of wages, adjusted with 1990's inflation rate. 
4. BLACK: binary variable, 1 indicates Black and 0 otherwise. 
5. HISPANIC: binary variable, 1 indicates Hispanic and 0 otherwise. 
6. HGC: the highest grade completed.
7. UERATE: the unemployment rate of the year of the survey. When missing, the variable is set to be 7.875 (the average rate).

@@

In the context of official statistics, @LooMarkvander2018Sdcw describe the "statistical value chain," which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables). What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain. In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).

@@

Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables. This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

@@

Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).

For `sex` and `race`, we have simplified the original names provided in the raw data, `SAMPLE_SEX_1979` and `SAMPLE_RACE_78SCRN`, respectively. However, it is important to note that the use of "sex" is probably not correct. This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in @SAGER. When using the dataset in the classroom, the educators might include discourse on the use of terms "sex" and "gender". Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification. From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.  @kennedy2020using provides a helpful discussion on these topics. <!--Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender. -->

@@

The next step is processing the highest grade completed. There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis. The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, encoded as a factor and integer variable, respectively. For each individual, there is only one value of `hgc` and `hgc_i`. This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on). We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types. Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.

The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as `grade`). This value can change over time, but it should only increase. This has been included in the refreshed dataset to enable richer downstream analyses. For example, if one would like to explore how the temporal changes in education level affect the wages of individuals. Hence, different from `hgc`, this variable could be used to demonstrate longitudinal and time series analysis in the classroom. This is recorded in `new_data_qnames` as columns beginning with `Q3-4` and `HGC` with year as a suffix. In addition, it is also in the columns beginning with `HGCREV` reflecting revised data. We chose to use these revised values because there were fewer missing values, indicating it had been more thoroughly checked. In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly. <!--We also check the consistency between `hgc_i` and `grade`. Some problematic observations were found, for example, `grade` value is greater than `hgc_i`. We then fix this problem by replacing those `grade` that is greater than its `hgc_i`. -->

@@

Our target variables to extract for employment are the respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (years) in workforce (`yr_wforce`), and work experience measured as the number of years worked (`exp`). As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.

@@

For the `stwork` variable,  we only rename the column names from `new_data_qnames`. This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual started working (`stwork`). Finally, the `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`. To obtain the work experience since 1979, we calculate the cumulative value. As the measurement unit is weeks, we convert this to year.

The employment and demographic variables are then joined. These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in @SingerJudithD2003Alda. We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data. However, it is worth noting that the original data does not restrict the number of observations for each individual, i.e., there are individuals with only one and two observations.

Finally, we save the resultant wage data on this cohort as `wages`. Note that we save the `grade` variable in this dataset instead of the `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.

@@

Experience is years after entering the labor force. It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`. However, using this calculation produced numbers that do not quite match the original data.

@@

In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides. The first step is validating that the numerical summaries of the raw data are the same as reported by NLSY79. This is followed by graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].

@@

Figure \ref{fig:feature-plot} shows an alternative way to check the data quality. Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals, mostly around 1990. Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals. This tells us there are a number of individuals with unbelievably high maximum wages. Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour. This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.

@@

The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers. It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important to wage volatility, such as abnormally high wages for a fleeting amount of time. To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed. Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined. Overly smooth profiles would indicate the replacement of values was too severe, resulting in removing the very interesting volatility of wages seen in many individuals. This was achieved using a purpose written `shiny`  [@shiny] app (provided with the code of this paper). <!--A threshold of 0.12 was chosen. That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.--> Observations where a change was made are tracked with a new variable called `is_pred`, so that this effect in the downstream analyses can be monitored.

@@

Figure \ref{fig:fixed-feature-plot} shows the summary statistics after removing extremes. The highest wage overall is now around \$1000. Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there has been a steady increase in wages over the years. Plot (B) shows there are still a small number of individuals with high maximum wages. Plot (C) shows the profile for ID=39 after imputing the extreme value. The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.

@@

  1. `id`: A unique individual's ID number. This is the `key` of the data as we saved the data as a `tsibble` object. 
  2. `year`: The year the observation was taken. This is the `index` of the data.
  3. `wage`: The mean of the hourly wages the individual gets at each of their different jobs. The value could be a weighted or an arithmetic mean. The weighted mean is used when the information of hours of work as the weight is available. The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
  4. `age_1979`: The age of the individual in 1979.
  5. `sex`: Sex of the individual, f = Female and m = Male.
  6. `race`: Race the individual belongs to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
  7. `grade`: Integer value of the highest grade completed corresponding to the survey year. 
  8. `hgc`: The highest grade completed ever.
  9. `hgc_i`: Integer value of the highest grade completed ever.
  10. `hgc_1979`: The highest grade completed in 1979 (integer value).

@@

1. `r nrow(join_more_17)` individuals were more than 17 years old in 1979. Thus, it looks like the description of the original data, that there are people older than 17 in the subset, is not quite accurate. Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later. 
2. `r nrow(join_less_17)` individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
    i. `r nrow(join_not_elig)` of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
    ii. The information about `ged` is missing in `r nrow(join_missing_ged)` individuals. We decided to include them in the refreshed data. 
    iii. `r nrow(join_ged_both)` individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
    iv. `r nrow(join_not_exist)` individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.

@@

```{r compare-subsets, fig.cap = "Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages. Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE}
```

@@

The variables from "highest grade completed" have some confusion. There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year. To match the original data, it is appropriate to use the `hgc` while in high school. The documentation suggests this variable is available, but it is not actually present in the data. Hence, to match the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value. The result does not exactly match the original for a few individuals.

In the original dataset, wages were inflation-adjusted to 1990 prices. This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey. Instead, we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyze the data.

@@

Ultimately, the refreshed data is reasonably similar to the original but unsatisfactorily far from it. The last step required would be to inflation-adjust wages. This is better to do with each wave of new data added so that it is relative to the last date in the data. Our decision was to provide the raw wages and include code to make the adjustment as part of the package.

@@

- It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges. Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
- Some of the methods shown, and lessons learned about the data, can also be considered when working with other datasets, and more generally for data cleaning, IDA, and EDA.
- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current for today's students.
- It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience. The vignettes of the `brolgar` [@brolgar] package provide useful guides for the case study content.
